manuscript

### 2024-01-23 manuscript/ms.Rmd
First completed figure/text draft

@@

Estimators of population genetic diversity are themselves extremely diverse, but can essentially be split into two categories depending on if they are allele frequency dependent or independent. Allele frequency dependent statistics, such as expected heterozygosity ($H_{e}$), the F-statistics ($F_{IS}$, $F_{ST}$, and $F_{IT}$)[@Weir1984], nucleotide diversity ($\pi$), and Tajima's $\theta$ [@Tajima1989] all rely principally on the frequency of the alleles across any surveyed loci; whereas allele frequency independent statistics, such as Watterson's $\theta$ [@wattersonNumberSegregatingSites1975], counts of alleles per loci, the number of observed segregating sites, and (to some degree) private allele counts do not. This is not a trivial difference:  evolutionary forces do not act equally on the number of loci/alleles in a population and the frequency of those alleles. Tajima's D, for example, is a powerful and broadly used statistic fundamentally based on the difference between Tajima's $\theta$ and Watterson's $\theta$ can be used to both detect selection and population demographic changes [@Tajima1989] due to the way that those forces act on the balance of allele frequencies within populations.

The average allele count per locus (which is usually corrected to allelic richness [@kalinowskiCountingAllelesRarefaction2004]) has historically been one of the most prevalent allele frequency independent measures of genetic diversity. While this is a particularly useful measure in microsatellite and other datasets utilizing heavily polyallelic markers, it is less so in single-nucleotide polymorphism (SNP) datasets. SNP datasets are usually primarily biallelic, and thus the count of the number of alleles per locus in such datasets is less variable across populations and therefore less interesting. Counts of the number of segregating sites per population are a useful alternative in SNP datasets given the large number of possible segregating sites that these datasets often query [@?CITE].

However, raw counts of the number of segregating sites per population are problematic whenever sample sizes are not equal across the populations under comparison at all loci. This can occur either directly due to unequal numbers of individuals or  indirectly due to unqual proportions of missing data across populations. This problem also effects estimates of allele counts per locus and private allele numbers, but corrections for both which use rarefaction to estimate those parameters under a common sample size are well developed [@kalinowskiCountingAllelesRarefaction2004].

@@

# Discussion
The number of segregating sites can provide an excellent counterpoint to measures of genetic diversity that are based on allele frequencies (such as observed and expected heterozygosity). For example, populations that have experienced a recent population expansion will often carry an excess of low frequency variants [@gattepailleInferringPopulationSize2013] caused by the recent increase in the overall rate at which mutations are produced in the population and the relative lack of time for any such new variants to drift to higher frequencies. The average expected heterozygosity across segregating sites may actually be lower (or at least lower than expected) in such cases than populations which have been demographically static, but they will carry far more segregating sites. A recent study in yellow perch provides an excellent example of such: several recently expanded populations show a relatively slight difference in heterozygosity in comparison to other, more demographically static populations but are segregating at far more loci [@schraidtDispersiveCurrentsExplain2023].
In cases like this, it is particularly important that the number of segregating sites is being properly calculated across populations, since failing to correct for differences in sample size or data missingness could mask biologically interesting signals of demographic history and obscure a critical facet of overall genomic diversity. The method we present here should therefore be a useful for future studies of genetic diversity.

@@

>>>>

Population geneticists use a wide range of approaches to measure population genetic diversity. While measures which incorporate allele frequencies (such as observed and expected heterozygosity, nucleotide diversity, Tajima's D, and so on) are well developed for single nucleotide polymorphism (SNP) data, there are only a few allele frequency independent estimators which are broadly in use. The number of segregating sites (or SNPs) is one of the more prevalent of these; however, unlike allelic richness or private allele frequencies, segregating site counts are rarely adjusted to correct for unequal sample sizes or missing data rates between populations. We here propose a corrected estimator for the number of segregating sites expected in a population following rarefaction, which allows for an unbiased approach for comparing the number of segregating sites between multiple populations.

@@

Estimators of population genetic diversity are themselves extremely diverse, but essentially fall into two categories: those that are depend on allele frequencies and those that do not. Allele frequency dependent statistics, such as expected heterozygosity ($H_{e}$), the F-statistics ($F_{IS}$, $F_{ST}$, and $F_{IT}$)[@Weir1984], nucleotide diversity ($\pi$), and Tajima's $\theta$ [@Tajima1989] all rely principally on the frequency of the alleles across any surveyed loci; whereas allele frequency independent statistics, such as Watterson's $\theta$ [@wattersonNumberSegregatingSites1975], counts of alleles per loci, the number of observed segregating sites, and private allele counts do not. This is not a trivial difference:  evolutionary forces do not act equally on the number of loci/alleles in a population and the frequency of those alleles. Tajima's D, for example, is a powerful and broadly used statistic fundamentally based on the difference between Tajima's $\theta$ and Watterson's $\theta$ that can be used to both detect selection and population demographic changes [@Tajima1989] due to the way that those forces act on the balance of allele frequencies within populations.

The average allele count per locus (which is usually corrected to allelic richness [@kalinowskiCountingAllelesRarefaction2004]) has historically been one of the most prevalent allele frequency independent measures of genetic diversity. While this is a particularly useful measure in microsatellite and other datasets utilizing heavily polyallelic markers, it is less so in single-nucleotide polymorphism (SNP) datasets. SNP datasets are usually biallelic by design, and thus allele counts per locus vary little populations and are therefore less informative. The number of segregating sites per population is a useful alternative in SNP datasets given the large number of possible segregating sites that these datasets often query [@?CITE].

However, segregating sites counts per population are problematic whenever sample sizes are not equal across the populations under comparison at all loci. This can occur either directly due to unequal numbers of individuals or  indirectly due to unqual proportions of missing data across populations. This problem also effects estimates of allele counts per locus and private allele numbers, but corrections for both which use rarefaction to estimate those parameters under a common sample size are well developed [@kalinowskiCountingAllelesRarefaction2004].

@@

# Discussion
We present here an method to correct for the probability that a given loci would be observed as segregating following rarefaction to a given sample size. This provides for the straightforward estimation of the total number of segregating sites which would be observed in a population sample at any reduced size, and therefore provides a way to standardize that metric across samples from different populations. We also show that this approach is unbiased by allele frequency or missing data variation across individual samples or loci.

Our estimator should useful given that the number of segregating sites can provide an excellent counterpoint to measures of genetic diversity that are based on allele frequencies (such as observed and expected heterozygosity). For example, populations that have experienced a recent population expansion will often carry an excess of low frequency variants [@gattepailleInferringPopulationSize2013] caused by the recent increase in the overall rate at which mutations are produced in the population and the relative lack of time for any such new variants to drift to higher frequencies. The average expected heterozygosity across segregating sites may actually be lower (or at least lower than expected) in such cases than populations which have been demographically static, but they will carry far more segregating sites. A recent study in yellow perch provides an excellent example of such: several recently expanded populations show a relatively slight difference in heterozygosity in comparison to other, more demographically static populations but are segregating at far more loci [@schraidtDispersiveCurrentsExplain2023].

In cases like this, it is particularly important that the number of segregating sites is being properly calculated across populations, since failing to correct for differences in sample size or data missingness could mask biologically interesting signals of demographic history and obscure a critical facet of overall genomic diversity. The $E(N_{S})$ method we present here should therefore be a useful for future studies of genetic diversity, and is currently available for use via the function "calc_seg_sites()" in the R package "snpR" [@hemstromSnpRUserFriendly2023].

@@

# Figure Captions:

Figure 1: The expected probabilities $P({S_{j,a}})$ of observing a segregating site at a locus for loci with different minor allele frequencies. Probabilities are derived from Equation 4 for each locus *a* for population sizes of 100 (left) and 1000 (right), each rarefacted to $n = 30$. Each locus has a random, independent percentage of missing data which varies between 0 and 30. Points are colored depending on if they fell within a 95% confidence level based on 10,000 random re-samplings of each population to 30 individuals.

Figure 2: The expected total number of segregating sites $E(N_{S})$ from populations of either 100 or 1,000 sampled to 30 individuals as calculated via Equation 5 vs that observed from 10,000 random re-samplings to 30 individuals. 95% Confidence and prediction intervals for $E(N_{S})$ are shown, calculated according to Equations 8 and 9. Observed $N_{S}$ values from each re-sample are colored depending on if they fell within the estimated 95% prediction interval. The mean $N_{S}$ across all re-samples is marked with a triangle, colored depending on if it fell within the estimated 95% confidence interval.
